---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# biometryassist

```{r, echo = FALSE}
description <- read.dcf('DESCRIPTION')
version <- as.vector(description[, 'Version'])
min.r <- substr(description[,"Depends"], 7, 11)
```

<!-- badges: start -->
[![Project Status: Active:  The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Codecov test coverage](https://codecov.io/gh/biometryhub/biometryassist/branch/main/graph/badge.svg)](https://app.codecov.io/gh/biometryhub/biometryassist?branch=main)
[![R build status](https://github.com/biometryhub/biometryassist/workflows/R-CMD-check/badge.svg)](https://github.com/biometryhub/biometryassist/actions)
[![minimal R version](https://img.shields.io/badge/R%3E%3D-`r min.r`-6666ff.svg)](https://cran.r-project.org/)
[![packageversion](https://img.shields.io/badge/Package%20version-`r gsub('-', '--', version)`-orange.svg?style=flat-square)](https://github.com/biometryhub/biometryassist/commits/main)
[![Licence](https://img.shields.io/github/license/mashape/apistatus.svg)](https://choosealicense.com/licenses/mit/)
[![Hits](https://hits.seeyoufarm.com/api/count/incr/badge.svg?url=https%3A%2F%2Fgithub.com%2Fbiometryhub%2Fbiometryassist&count_bg=%2379C83D&title_bg=%23555555&icon=&icon_color=%23E7E7E7&title=hits&edge_flat=false)](https://hits.seeyoufarm.com)
<!-- badges: end -->


The goal of biometryassist is to provide functions to aid in the Design and Analysis of Agronomic-style experiments through easy access to documentation and helper functions, especially while teaching these concepts.

*This package is a renamed version of BiometryTraining which is no longer maintained, but can still be found at https://biometryhub.github.io/BiometryTraining/*

----

## Installation

Use the following code to install this package:

```{r eval=F}
if(!require("remotes")) install.packages("remotes") 
remotes::install_github("biometryhub/biometryassist", upgrade = FALSE)
```


## Using the package

Load the package and start using it with:

```{r eval=F}
library(biometryassist)
```

If you find this package useful, please cite it! Type `citation("biometryassist")` on the R console to find out how.

## Troubleshooting Installation

- If you have the latest version already installed, you will get a response like:

<span style="color: red;">Skipping install of 'biometryassist' from a github remote, the SHA1 (e034b603) has not changed since last install.<br>
  Use `force = TRUE` to force installation</span>

- The following warning messages can usually be ignored:

<span style="color: red;">WARNING: Rtools is required to build R packages, but is not currently installed.</span>

<span style="color: red;">Please download and install Rtools custom from https://cran.r-project.org/bin/windows/Rtools/.</span>

<span style="color: red;">Warning messages:<br>
1: In untar2(tarfile, files, list, exdir) :<br>
  skipping pax global extended headers<br>
2: In untar2(tarfile, files, list, exdir) :<br>
  skipping pax global extended headers</span>

- If you receive an error that the package could not install because `rlang` or another package could not be upgraded, the easiest way to deal with this is to uninstall the package(s) that could not be updated (`remove.packages("rlang")`). Then re-install with `install.packages("rlang")` and then try installing `biometryassist` again.

