---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# biometryassist

```{r, echo = FALSE}
description <- read.dcf('DESCRIPTION')
version <- as.vector(description[, 'Version'])
min.r <- substr(description[,"Depends"], 7, 11)
```

<!-- badges: start -->
[![Project Status: Active:  The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Codecov test coverage](https://codecov.io/gh/biometryhub/biometryassist/branch/main/graph/badge.svg)](https://app.codecov.io/gh/biometryhub/biometryassist?branch=main)
[![R build status](https://github.com/biometryhub/biometryassist/workflows/R-CMD-check/badge.svg)](https://github.com/biometryhub/biometryassist/actions)
[![minimal R version](https://img.shields.io/badge/R%3E%3D-`r min.r`-6666ff.svg)](https://cran.r-project.org/)
[![packageversion](https://img.shields.io/badge/Package%20version-`r gsub('-', '--', version)`-orange.svg?style=flat-square)](https://github.com/biometryhub/biometryassist/commits/main)
[![Licence](https://img.shields.io/github/license/mashape/apistatus.svg)](https://choosealicense.com/licenses/mit/)
[![](https://cranlogs.r-pkg.org/badges/biometryassist)](https://cran.r-project.org/package=biometryassist)
![Badge](https://hitscounter.dev/api/hit?url=https%3A%2F%2Fbiometryhub.github.io%2Fbiometryassist&label=Hits&icon=arrow-down-circle&color=%23198754)
<!-- badges: end -->


The goal of biometryassist is to provide functions to aid in the Design and Analysis of Agronomic-style experiments through easy access to documentation and helper functions, especially while teaching these concepts.

*This package is a renamed version of BiometryTraining which is no longer maintained, but can still be found at https://biometryhub.github.io/BiometryTraining/*

----

## Installation

As of version 1.0.0 the biometryassist package is now [on CRAN](https://cran.r-project.org/package=biometryassist) 🙌
That means that installation is as easy as running:

```{r installing_cran, eval=FALSE}
install.packages("biometryassist")
```


### Development version

⚠ **Warning**: The development version is unstable and liable to change more often than the CRAN version. It may have bugs fixed, but there may be other currently unknown bugs introduced. ⚠


Use the following code to install the latest development version of this package.

```{r eval=F}
if(!require("pak")) install.packages("pak")
pak::pak("biometryhub/biometryassist@dev")

# Alternatively
if(!require("remotes")) install.packages("remotes")
remotes::install_github("biometryhub/biometryassist@dev")
```


## Using the package

Load the package and start using it with:

```{r eval=F}
library(biometryassist)
```

The package supports two main workflows:

**Experimental design** — generate and visualise trial layouts:

- `design()` — create CRD, RCBD, Latin Square, split-plot, strip-plot, and factorial designs, with a plot of the layout and a skeletal ANOVA table
- `add_buffers()` — add buffer plots around treatment plots, blocks, or the trial perimeter
- `export_design_to_excel()` — write a design to a formatted Excel workbook

**Post-model analysis and visualisation** — works with models from `aov`, `lme4`, `nlme`, `asreml`, `sommer`, `glmmTMB`, `afex`, and more:

- `multiple_comparisons()` — Tukey HSD and other multiple comparison tests with letter groupings
- `pairwise_comparisons()` — selected pairwise differences or general linear contrasts
- `reference_comparisons()` — compare all levels against a reference (Dunnett-style)
- `resplot()` — diagnostic residual plots
- `heat_map()` — spatial heat maps of trial data or residuals
- `variogram()` — spatial variograms for ASReml-R models

The package optionally enhances the commercial [ASReml-R](https://vsni.co.uk/software/asreml-r) package. If you have a licence, `install_asreml()` and `update_asreml()` make it easy to install and keep up to date.

For details on recent changes, see [NEWS.md](NEWS.md).

### Example

```{r example, message=FALSE}
library(biometryassist)

# Generate a Randomised Complete Block Design with 11 treatments and 4 reps
des.out <- design(
    type = "rcbd",
    treatments = LETTERS[1:11],
    reps = 4,
    nrows = 11,
    ncols = 4,
    brows = 11,
    bcols = 1,
    seed = 42,
    quiet = TRUE
)
```

```{r example-plot, echo=FALSE}
autoplot(des.out)
```

The `$satab` element gives the skeletal ANOVA table for the design:

```{r example-satab}
cat(des.out$satab)
```


## Troubleshooting Installation

- If you receive an error that the package could not install because `rlang` or another package could not be upgraded, the easiest way to deal with this is to uninstall the package(s) that could not be updated (`remove.packages("rlang")`). Then restart R, re-install with `install.packages("rlang")` and then try installing `biometryassist` again.


## Citation

If you find this package useful in your work, please cite it. Run the following in R to get the citation details:

```{r citation, eval = FALSE}
citation("biometryassist")
```

```{r citation_run, echo=FALSE, warning=FALSE}
# CRAN populates the publication date automatically, so DESCRIPTION
# deliberately has no Date field. Without one, citation() can't determine a
# year and warns, printing "(????)". To render a tidy citation here, copy the
# installed DESCRIPTION into a temporary library with the current date added
# and cite from there, leaving the committed DESCRIPTION untouched.
local({
  tmp_lib <- file.path(tempdir(), "cite_lib")
  pkg_dir <- file.path(tmp_lib, "biometryassist")
  dir.create(pkg_dir, recursive = TRUE, showWarnings = FALSE)
  desc <- readLines(system.file("DESCRIPTION", package = "biometryassist"))
  writeLines(
    c(desc, paste0("Date: ", Sys.Date())),
    file.path(pkg_dir, "DESCRIPTION")
  )
  citation("biometryassist", lib.loc = tmp_lib)
})
```

## Code of Conduct

Please note that the biometryassist project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.


## Contributing

Contributions are welcome! Please see [CONTRIBUTING.md](CONTRIBUTING.md) for guidelines on how to get involved.


## Licence

MIT © University of Adelaide Biometry Hub. See [LICENSE.md](LICENSE.md) for details.
