---
title: "Choosing and interpreting multiple comparisons"
format: html
vignette: >
  %\VignetteIndexEntry{Choosing and interpreting multiple comparisons}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
engine: knitr
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(biometryassist)
```

## Introduction

After fitting a model, a common next step is to compare predicted treatment
means. `biometryassist` provides three complementary functions for this, and
which one you want depends on *what you are comparing*:

- **`multiple_comparisons()`** is **means-centric**. It returns one row per
  treatment level (mean, standard error, confidence interval) and summarises the
  *complete* set of all pairwise comparisons with a compact letter display
  (treatments that share a letter show no evidence of a difference).
- **`pairwise_comparisons()`** is **difference-centric**. It returns a tidy
  table with one row per requested comparison (estimate, standard error,
  statistic, degrees of freedom and an adjusted p-value), and is the honest
  representation when you only care about *some* of the comparisons.
- **`reference_comparisons()`** is means-centric, for the specific question
  "how does each treatment compare to a single control?". It returns each
  level's mean, the reference mean and the adjusted difference, using an exact
  Dunnett test by default.

This vignette introduces `pairwise_comparisons()` and `reference_comparisons()`,
shows them alongside `multiple_comparisons()`, and then discusses how to choose a
multiplicity adjustment and which function fits your question. All examples use
built-in data and `aov()` so they can be run without any specialised modelling
packages, but all three functions work across the supported model types (`aov`,
`lm`, `aovlist`, `asreml`, `lme4`, `nlme`).

## A quick tour of `pairwise_comparisons()`

We will use the `iris` data and a one-way model for petal width.

```{r}
model <- aov(Petal.Width ~ Species, data = iris)
```

By default `pairwise_comparisons()` tests **all** pairwise differences and
adjusts the p-values with Holm's method:

```{r}
pairwise_comparisons(model, classify = "Species")
```

Each row is a difference: `estimate` is `mean(level1) - mean(level2)`, with
`level1`/`level2` making the direction explicit, and the `level1.mean` /
`level2.mean` columns reporting the two predicted means it is computed from. If
you only care about a
specific set of comparisons, pass them via `pairs` (the multiplicity adjustment
is then applied over just that set):

```{r}
pairwise_comparisons(
  model,
  classify = "Species",
  pairs = c("setosa-versicolor", "versicolor-virginica")
)
```

`autoplot()` draws a forest plot of the differences, with a dashed line at zero
and an asterisk marking comparisons that are significant at the (adjusted) level:

```{r, fig.height = 3}
autoplot(pairwise_comparisons(model, classify = "Species"))
```

### General contrasts

Sometimes the comparison of interest isn't a single pair. The `contrasts`
argument tests any linear contrast — for example, *setosa* versus the average of
the other two species:

```{r}
pairwise_comparisons(
  model,
  classify = "Species",
  contrasts = list(
    "setosa vs rest" = c(setosa = 1, versicolor = -0.5, virginica = -0.5)
  )
)
```

Each contrast is a named vector of coefficients (which should sum to zero) and
the list name becomes the row label. A two-level contrast such as
`c(A = 1, B = -1)` is exactly the pairwise difference `"A-B"`.

### The same model, two views

Contrast that with `multiple_comparisons()` on the same model, which reports the
means and their letter groupings:

```{r}
multiple_comparisons(model, classify = "Species")
```

Both are correct; they answer different questions. The letter display is a
compact summary of *every* pairwise comparison at once — its contract is that
all pairs have been compared, so "share a letter" means "no evidence of a
difference". That summary is only meaningful for the **complete** set: with an
incomplete or irregular set of comparisons there is no valid letter display, and
forcing one would silently assert "not different" for pairs that were never
tested. When you are interested in a chosen subset of comparisons, the tidy
difference table is the honest representation.

## Controlling error across comparisons

Testing several comparisons inflates the chance of at least one false positive,
so p-values are adjusted. There are two broad families of error rate, and the
right choice depends on your goal.

**Family-wise error rate (FWER)** is the probability of making *any* false
rejection across the whole family. It is the conservative, "every claim must be
trustworthy" criterion, appropriate for a small set of planned comparisons.

- **Bonferroni** is the simplest FWER method but is unnecessarily conservative.
- **Holm** [Holm, 1979] is a step-down procedure that controls the FWER under
  *any* dependence structure and is uniformly more powerful than Bonferroni —
  there is essentially never a reason to prefer plain Bonferroni over Holm. It
  is the default in `pairwise_comparisons()`.
- **Tukey's HSD** [Tukey, 1953; Hsu, 1996] uses the studentized range
  distribution to give *exact* simultaneous FWER control for the complete set of
  **all** pairwise comparisons (exact for equal replication; the Tukey–Kramer
  extension covers unequal replication). Because it exploits the exact joint
  distribution of all pairwise differences, it is more powerful than Bonferroni
  or Holm *for that complete set* — which is why it is the default in
  `multiple_comparisons()`.
- **Dunnett's test** [Dunnett, 1955] is the exact FWER procedure for the
  specific family of comparing every treatment against a *single control*. Like
  Tukey for all pairs, it exploits the known joint distribution of those
  comparisons (the correlation induced by the shared control), so it is more
  powerful than a generic adjustment for that family. It is the default in
  `reference_comparisons()`.

**False discovery rate (FDR)** is the expected *proportion* of false rejections
among the rejected comparisons. It is less stringent than FWER and earns its
keep when comparisons are many and somewhat exploratory.

- **Benjamini–Hochberg (`"BH"`/`"fdr"`)** [Benjamini & Hochberg, 1995] controls
  the FDR under independence and positive dependence (which pairwise comparisons
  satisfy); **`"BY"`** is valid under arbitrary dependence but more
  conservative.

Both functions accept any method supported by `stats::p.adjust()` through their
`adjust` argument, so you can choose the criterion that matches your goal.

### Why Tukey is the means function's job

A natural question is why `pairwise_comparisons()` does not offer
`adjust = "tukey"`. Tukey's critical value is calibrated to the largest
difference among *all* the means; applied to a chosen subset it still controls
the FWER, but it is sized for comparisons you are not making, so it is
conservative and conceptually mismatched [Hsu, 1996; Bretz et al., 2010]. For a
selected set, a method calibrated to *that* set (Holm, Bonferroni, BH) is
preferred. Conversely, for the complete set Tukey is the most powerful of these
because it uses the known structure of the pairwise differences. So Tukey
belongs with the complete-set, means-and-letters analysis
(`multiple_comparisons()`), and `pairwise_comparisons()` rejects it with a
message pointing you there.

Because both functions share the same predicted means and standard errors of
differences, **testing all pairs in `pairwise_comparisons()` with a given
`adjust` reproduces the adjusted p-values of `multiple_comparisons()` with the
same `adjust`** — only the presentation differs (a table of differences versus
means and letters). The one exception is Tukey, for the reasons above.

## Comparing each treatment to a control

When the question is specifically "how does each treatment compare to a single
control?", `reference_comparisons()` is the natural tool. It is means-centric
like `multiple_comparisons()`, but only the comparisons against the chosen
reference are made — and because every comparison shares that reference,
significance attaches cleanly to each treatment's mean (which a letter display
cannot do for an incomplete set).

Using the `chickwts` data (chick weights by feed type), comparing every feed to
the `casein` control:

```{r}
model_cw <- aov(weight ~ feed, data = chickwts)
reference_comparisons(model_cw, classify = "feed", reference = "casein")
```

Each row gives a feed's mean (`level1.mean`), the control mean (`level2.mean`)
and their difference, with the adjustment defaulting to the exact **Dunnett**
test. The `autoplot()` method draws each feed's mean, a dashed line at the
control mean (marked with a diamond), and an interval for the difference from the
control, so the interval clears the control line exactly when the comparison is
significant; significant feeds are also flagged with an asterisk:

```{r, fig.height = 3.5}
autoplot(reference_comparisons(model_cw, classify = "feed", reference = "casein"))
```

As with the other functions, `by` runs the comparisons independently within
subgroups — for example comparing each treatment to the control *within each
site or year*. (Dunnett needs a single residual degrees-of-freedom; for models
that report comparison-specific df, `reference_comparisons()` falls back to Holm
with a message.)

## The `adjust` and `by` arguments of `multiple_comparisons()`

`multiple_comparisons()` also accepts `adjust` for choosing the p-value
adjustment method. The default is Tukey's HSD; any `stats::p.adjust()` method can
be used instead:

```{r}
mc <- multiple_comparisons(model, classify = "Species", adjust = "bonferroni")
mc$comparison_method
```

The `by` argument runs comparisons **independently within subgroups** — useful
for a factorial where you want to compare one factor *within* each level of
another, rather than across the whole experiment. Each subgroup is a separate
family, with no pooling or adjustment across groups. Using the `warpbreaks`
data (number of breaks by `wool` type and `tension`):

```{r}
wb <- aov(breaks ~ wool * tension, data = warpbreaks)
```

Comparing `tension` levels within each `wool` type, the comparison labels refer
to the remaining (non-`by`) factor:

```{r}
pairwise_comparisons(wb, classify = "wool:tension", by = "wool")
```

`by` works the same way in `multiple_comparisons()`, where `autoplot()` then
facets the means and letters by the `by` variable:

```{r, fig.height = 3.5}
autoplot(multiple_comparisons(wb, classify = "wool:tension", by = "wool"))
```

## A note on confidence intervals

The intervals shown by `pairwise_comparisons()` (and its forest plot) are
*per-comparison* intervals — they are **not** adjusted for simultaneity.
Significance, on the other hand, is judged by the multiplicity-adjusted
p-value. These two can disagree: an interval may exclude zero while the adjusted
p-value is no longer below `sig`. This is the same tension that the `int.type`
options address in `multiple_comparisons()`. So in the forest plot, read
significance from the asterisk (which tracks the adjusted test), not from
whether the drawn interval happens to cross zero.

## Choosing: a short guide

- **All pairwise comparisons, with a compact summary?** Use
  `multiple_comparisons()` (Tukey's HSD by default, with letter groupings).
- **A specific or planned subset of comparisons?** Use
  `pairwise_comparisons()` (Holm by default), which represents the chosen set
  honestly as a table of differences.
- **Each treatment against a single control?** Use `reference_comparisons()`
  (exact Dunnett by default), which keeps the analysis means-centric.
- **Many, exploratory comparisons where a controlled proportion of false
  positives is acceptable?** Use an FDR method (`adjust = "BH"`).
- **Comparisons within subgroups of a factorial?** Use `by` in either function.

## References

- Benjamini, Y., & Hochberg, Y. (1995). Controlling the false discovery rate: a
  practical and powerful approach to multiple testing. *Journal of the Royal
  Statistical Society, Series B (Methodological)*, 57(1), 289–300.
  <doi:10.1111/j.2517-6161.1995.tb02031.x>
- Bretz, F., Hothorn, T., & Westfall, P. (2010). *Multiple Comparisons Using R*.
  Boca Raton: Chapman & Hall/CRC.
- Dunnett, C. W. (1955). A multiple comparison procedure for comparing several
  treatments with a control. *Journal of the American Statistical Association*,
  50(272), 1096–1121. <doi:10.1080/01621459.1955.10501294>
- Holm, S. (1979). A simple sequentially rejective multiple test procedure.
  *Scandinavian Journal of Statistics*, 6(2), 65–70.
- Hothorn, T., Bretz, F., & Westfall, P. (2008). Simultaneous inference in
  general parametric models. *Biometrical Journal*, 50(3), 346–363.
  <doi:10.1002/bimj.200810425>
- Hsu, J. C. (1996). *Multiple Comparisons: Theory and Methods*. London:
  Chapman & Hall.
- Miller, R. G. (1981). *Simultaneous Statistical Inference* (2nd ed.). New York:
  Springer-Verlag.
- Tukey, J. W. (1953). The problem of multiple comparisons. Unpublished
  manuscript. Reprinted in H. I. Braun (Ed.) (1994), *The Collected Works of
  John W. Tukey, Volume VIII: Multiple Comparisons, 1948–1983* (pp. 1–300). New
  York: Chapman & Hall.
```
