---
title: "Installing ASReml-R"
output: 
    rmarkdown::html_vignette:
        toc: yes
vignette: >
  %\VignetteIndexEntry{Installing ASReml-R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

<details>
<summary style="font-size: 125%; font-weight: 700;">
Quick Version
</summary>
Run the following code to install and licence ASReml-R:

```{r}
install.packages("biometryassist")
biometryassist::install_asreml()
asreml::asreml.license.activate()
```

Then paste licence code at the prompt.

</details>

## Package method

The easiest way to install ASReml-R is to use the `biometryassist` package, which includes an automated `install_asreml()` function. This function will detect the appropriate version of R and your operating system, and download the matching version of asreml. If you'd rather not install the full package for this one function, you can simply source the function from the public GitHub repository instead, and run the function. See [Function Only Method].

The `biometryassist` package is on CRAN, so can be installed using the regular method:

```{r}
install.packages("biometryassist")
```

Once installed, load the package ready for use with:

```{r}
library(biometryassist)
```

This package will stay installed between R sessions, but will need to be loaded with the library command any time you want to use the functions within it.

## Function Only Method

To use only the function, without the additional overhead of the biometryassist package, source it with the following command:

```{r}
source("https://raw.githubusercontent.com/biometryhub/biometryassist/main/R/install_asreml.R")
```

This will enable use of the function within the current R session. If you close R, you will need to run the line above again to enable use of the function.

## Installing ASReml-R

Once the install function is available, simply run it and wait for it to finish:

```{r}
install_asreml()
```

The following output is typical:

```r
> install_asreml()

Downloading and installing ASReml-R. This may take some time, depending on internet speed...

* installing *source* package ‘asreml’ ...
** using staged installation
** R
** data
*** moving datasets to lazyload DB
** inst
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
** building package indices
** testing if installed package can be loaded from temporary location
** checking absolute paths in shared objects and dynamic libraries
** testing if installed package can be loaded from final location
** testing if installed package keeps a record of temporary installation path
* DONE (asreml)
ASReml-R successfully installed!
```
## Licencing ASReml-R

ASReml-R requires a licence code to activate and run. This licence code is checked for validity via the internet, and may require opening some network ports on your computer to successfully activate.

To activate the licence code once asreml is installed, run the following commands (noting the spelling of licen**s**e):

```{r}
library(asreml)
asreml.license.activate()
```

You will then be prompted to provide a licence code, which is a 16 character alpha-numeric code:

```r
> asreml.license.activate()
Please enter your activation code (RET or 0 to exit): AAA1-BBB2-CCC3-DDD4
```

A valid licence code will produce the following output:

```
License activation successful
Licensed to <Orginisation or name> - <reference code(s)>
Ref: <reference code> Expires: <expiry date>, <n> days.
```

### Licence key for University of Adelaide Staff and Students

The University of Adelaide has unlimited licence codes for university staff, students and affiliates, with thanks to funding provided by the Grains Research and Development Corporation, via the Analytics for the Australian Grains Industry (AAGI) project.

To access the University of Adelaide licence key, please complete this form:
https://forms.office.com/r/gBva4iL3FU

You will receive an email with a licence key if you hold a valid University of Adelaide account.

## Additional Installation Options

The `install_asreml()` function has a number of additional arguments, which you can read about [on the documentation site](https://biometryhub.github.io/biometryassist/reference/install_asreml.html).

If you already have ASReml-R installed, and want to update it or reinstall it, you can use the `update_asreml()` function, or `install_asreml(force = TRUE)`.

```{r}
update_asreml()
install_asreml(force = TRUE) # Equivalent to the above
```

