---
title: "Diagnosing Model Fit with resplot()"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-depth: 3
    toc-location: right   # or "right", "body"
    code-fold: true       # collapsible code blocks
    code-tools: true      # copy button + show/hide all
    number-sections: true
    fig-width: 7
    fig-height: 5
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Diagnosing Model Fit with resplot()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
engine: knitr
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, message=FALSE}
library(biometryassist)
library(ggplot2)
library(patchwork)
```

## Introduction

After fitting a model to data, it is important to check the residuals — the leftover variation in the data to gain important insight. In particular, we want to check whether the fitted model is appropriate for the data and adheres to key modelling assumptions. These assumptions may include approximately normal and homogeneous residual variance, and linearity in the case of linear regression. In `biometryassist` library, `resplot()` is a function that provides a single-panel diagnostic plot for a fitted model.

A call to `resplot()` always returns four core panels, built from the model's
*standardised* residuals (raw residuals divided by their standard deviation):

- **A — Histogram** of the standardised residuals, to check the overall shape of
  the residual distribution.
- **B — Normal Q-Q plot**, to check whether the residuals follow a normal
  distribution, especially in the tails.
- **C — Residuals vs. fitted values**, to check for non-constant variance or
  remaining structure that the model hasn't captured.
- **D — A Shapiro-Wilk normality test**, displayed as text with relevant p-values and conclusion underneath the three plots (toggle with and `shapiro = TRUE` as the default).

```{r basic_example}
model <- aov(Petal.Length ~ Petal.Width, data = iris)
resplot(model)
```
<!-- Need Check -->
`resplot()` works across most of the modelling engines supported by
`biometryassist` — `aov`/`lm`, `nlme::lme()`, `lme4`/`lmerTest`, `ASReml-R`,
`sommer`, `ARTool` and others, by dispatching on the model's class via an internal
`extract_model_info()` generic. See the
[*Supported model types*](#supported-model-types) section below.

## Modelling Assumptions and the Panels

A linear (mixed) model has the general form:

$$ \mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{Z}\mathbf{u} + \boldsymbol{\varepsilon}, \qquad \boldsymbol{\varepsilon} \sim \mathcal{N}(0, \sigma^2 \mathbf{I}) $$

where $\mathbf{y}$ is the vector of $n$ observed responses, $\mathbf{X}$ is the $n \times p$ design matrix for the $p$ fixed effects $\boldsymbol{\beta}$, and $\mathbf{Z}\mathbf{u}$ plays the equivalent role for the random effects ($\mathbf{Z}$ the design matrix, $\mathbf{u}$ the vector of random effects) — a fixed-effects-only model such as `aov()`/`lm()` simply omits this term. What matters most for residual diagnostics is the last piece, the residual error vector $\boldsymbol{\varepsilon}$, which has one entry per observation. Its assumed distribution, $\mathcal{N}(0, \sigma^2\mathbf{I})$, encodes four assumptions that any linear (mixed) model relies on:

1. **Linearity**: The relationship between the response and the fixed covariates is linear (or has been made linear through an appropriate transformation).

2. **Independence of errors**: The errors are statistically independent of one another, as encoded by the identity matrix $\mathbf{I}$ — there are no off-diagonal covariance terms.

3. **Normality of errors**: The errors $\boldsymbol{\varepsilon}$ are normally distributed at every value of the fixed covariates, as encoded by $\mathcal{N}(\cdot)$.

4. **Equal variance (homoscedasticity)**: The errors have constant variance $\sigma^2$ regardless of the fitted value, as encoded by the single scalar $\sigma^2$ multiplying $\mathbf{I}$ rather than a value that changes per observation.

These are all statements about $\boldsymbol{\varepsilon}$, which we can't observe directly, rather, it depends on the true (unknown) $\boldsymbol{\beta}$ and $\mathbf{u}$. What we *can* compute, once a model is fitted, is what's left over: for observation $i$, the raw residual $r_i = y_i - \hat{y}_i$ is the difference between the observed response and the model's fitted value, and stands in for the unobservable $\varepsilon_i$. `resplot()`'s panels are all built from these residuals, which is how the four assumptions above get checked in practice.

### Table linking the panels to these assumptions

Each panel checks a different one of these assumptions:

| Assumption | Panel(s) | What to look for |
|---|---|---|
| **Linearity** | **C** — Residuals vs. fitted | Random scatter around zero; a curved trend signals a missing non-linear term |
| **Normality of errors** | **A** — Histogram, **B** — Normal Q-Q plot, **D** — Shapiro-Wilk test | A symmetric, bell-shaped histogram; points hugging the Q-Q reference line; a Shapiro-Wilk p-value > 0.05 |
| **Equal variance (homoscedasticity)** | **C** — Residuals vs. fitted | Constant vertical spread across fitted values; a funnel/fan shape signals non-constant variance |

**Independence of errors** isn't something any of `resplot()`'s panels can check directly — it depends on the spatial or temporal arrangement of the errors, not just their marginal distribution. For trials laid out in rows and columns, use `variogram()` (or `heat_map()`) instead to check for spatial autocorrelation left in the residuals.

To see each panel in action, the rest of this section uses three example model fits: the well-behaved `model_good` simulated from a normal distribution, plus two models (`model_het` and `model_skew`) built to violate a specific assumption on purpose. Each panel below is pulled directly from `resplot()`'s output for these models using `resplot(model)[[i]]`, with `i = 1, 2, 3` corresponding to the three plots.

```{r example_models_setup}
set.seed(44)
n <- 100
x <- runif(n, 1, 10)
y_good <- 5 + 2 * x + rnorm(n, sd = 1)
model_good <- lm(y_good ~ x)

set.seed(88)
n <- 100
x <- runif(n, 1, 10)
y_het <- (2 + 0.8 * x) * exp(rnorm(n, sd = 0.3))
model_het <- lm(y_het ~ x)

set.seed(99)
x <- runif(100, 1, 10)
y_skew <- exp(1 + 0.5 * x + rnorm(100, sd = 0.5))
model_skew <- lm(y_skew ~ x)
```

`model_het`'s errors are simulated with `sd = x * 0.3`, so their spread grows with `x`, violating equal variance, while `model_skew`'s errors are exponentiated, giving them a right-skewed distribution instead of a normal one, violating normality. This isn't just a technicality: the standard errors, confidence intervals, and t-/F-tests used to compare treatment means are all derived assuming these assumptions hold, so for models like these, the p-values and confidence intervals `resplot()` and downstream comparisons report can end up too liberal or too conservative to trust.

To keep comparisons on a common scale, `resplot()` standardises the raw residuals before plotting them,

$$ r_i^{\text{std}} = \frac{r_i}{\hat{\sigma}\sqrt{(n-1)/n}} $$

where $\hat\sigma$ is their sample standard deviation. If the model's assumptions hold, $r_i^{\text{std}}$ should behave like an independent random sample from $\mathcal{N}(0,1)$, which `model`'s residuals broadly do, unlike `model_het` or `model_skew` by design. Panels A–D below each check a different one of the assumptions, as summarised in the table above.

### A — Histogram

The histogram shows the distribution of the standardised residuals $r_i^{\text{std}}$.
Use it to judge the **overall shape** of the residual distribution against the bell-shaped $\mathcal{N}(0,1)$ density it should follow.

```{r panel_a_example, fig.width=7, fig.height=3, fig.align='center'}
p_hist_good <- resplot(model_good, shapiro = FALSE)[[1]] +
    labs(title = "Good: symmetric, bell-shaped")

p_hist_bad <- resplot(model_skew, shapiro = FALSE)[[1]] +
    labs(title = "Warning: right-skewed")

p_hist_good + p_hist_bad
```

A roughly symmetric, bell-shaped histogram centred on zero (left) is consistent with normally distributed residuals. Strong right or left skew, a bimodal distribution, or very heavy tails (right) suggest the model is missing something: like a transformation of the response, an omitted covariate, or an unaccounted-for grouping factor.

### B — Normal Q-Q Plot

The Q-Q plot compares the empirical quantiles of $r_i^{\text{std}}$ to the
theoretical quantiles of $N(0,1)$ — the most direct check of the normality half
of the assumption. If the residuals are normally distributed, the points should
fall close to the diagonal reference line.

```{r panel_b_example, fig.width=7, fig.height=3, fig.align='center'}
p_qq_good <- resplot(model_good, shapiro = FALSE)[[2]] +
    labs(title = "Good: points on the reference line")

p_qq_bad <- resplot(model_skew, shapiro = FALSE)[[2]] +
    labs(title = "Warning: S-curve (skew)")

p_qq_good + p_qq_bad
```

When residuals are well-behaved (left), the points hug the diagonal line. Common
warning patterns include:

- **S-shaped curve** (right): the residuals are skewed — one tail is longer than
  the other.
- **Both ends curling away from the line**: heavy tails — more extreme values than
  a normal distribution would produce.
- **A single isolated point far from the rest**: a potential outlier that may be
  distorting the fit.

### C — Residuals vs. Fitted Values

This scatterplot places $r_i^{\text{std}}$ on the vertical axis and the model's fitted values on the horizontal axis. It checks the constant-variance half of the assumption: whether $\text{Var}(r_i^{\text{std}})$ stays flat across fitted values, as $\sigma^2$ in $\boldsymbol{\varepsilon} \sim \mathcal{N}(0, \sigma^2 \mathbf{I})$ is assumed constant rather than a function of $\hat{y}_i$.

```{r panel_c_example, fig.width=7, fig.height=3, fig.align='center'}
p_rf_good <- resplot(model_good, shapiro = FALSE)[[3]] +
    labs(title = "Good: random scatter around zero")

p_rf_bad <- resplot(model_het, shapiro = FALSE)[[3]] +
    labs(title = "Warning: funnel (heteroskedasticity)")

p_rf_good + p_rf_bad
```

Ideally the points form a random cloud centred around zero with roughly constant spread across all fitted values (left). Common problems to watch for:

- **Funnel or fan shape** (right): residual variance increases (or decreases) with the fitted value — heteroskedasticity. A variance-stabilising transformation (e.g. log, square-root) or a weighted model may be needed.
- **Curved trend**: a polynomial or non-linear term is missing from the model.
- **Distinct bands or clusters**: a factor that should be in the model has been omitted or mislabelled.

### D — Shapiro-Wilk Test

Below the three plots, `resplot()` (by default) displays the result of a Shapiro-Wilk normality test on the raw residuals. The formal test consits:  $H_0: r_i^{\text{std}} \sim \mathcal{N}(0,1)$, complementing the visual evidence in panels A and B. The output always includes the test statistic and a
plain-language interpretation. When the test finds no evidence of non-normality (p > 0.05):

```
Shapiro-Wilk normality test p-value: 0.3241
The residuals appear to be normally distributed. (n = 48)
```

When the test detects significant non-normality (p ≤ 0.05):

```
Shapiro-Wilk normality test p-value: 0.0003
The residuals do not appear to be normally distributed. (n = 48)
```

Set `shapiro = FALSE` to omit it entirely:

```{r shapiro_off}
resplot(model_good, shapiro = FALSE)
```

The Shapiro-Wilk test becomes unreliable as a guide for larger samples, that is, with enough observations, even trivial, practically meaningless departures from normality become "significant". `resplot()` reflects this:

- For 2,000–4,999 observations, the test still runs but a warning is issued.
- For 5,000 or more observations, the test is skipped (with a warning) and panel D is omitted regardless of the `shapiro` argument.

In both cases, it is good practice to judge normality visually from panels A and B rather than relying on the test alone.

## Understanding the Full Outputs

### Example: Simulate Normal Data

Before looking at real data, it helps to see what `resplot()`'s output looks like when the modelling assumptions hold *by construction*. The example below simulates data with a genuinely linear relationship and independent, normally distributed, constant-variance errors, then fits a model to it:

```{r full_output_sim, fig.align='center'}
resplot(model_good)
```

Reading each panel in this output:

- **A (histogram)**: symmetric and bell-shaped, matching the $\mathcal{N}(0,1)$ density it was drawn from.
- **B (Q-Q plot)**: points sit tightly on the reference line, right out to the tails.
- **C (residuals vs. fitted)**: an even, random scatter around zero with no funnel, curve, or banding.
- **D (Shapiro-Wilk)**: a large p-value, well above 0.05, with nothing in the visual panels to contradict it.

Because the errors here were simulated directly from $\mathcal{N}(0,1)$, every panel lines up with the textbook picture almost exactly. Treat this as the benchmark for what "no violations" looks like — real data, as the next example shows, is rarely this clean.

### Example: Real Data (`iris`)

Fitting a simple one-way model to the built-in `iris` data gives residuals that are broadly, but not perfectly, consistent with the model assumptions, as shown below:

```{r basic_example_2, fig.align='center'}
resplot(model)
```

Reading each panel in this output:

- **A (histogram)**: roughly symmetric and bell-shaped, which is consistent with normally distributed residuals.
- **B (Q-Q plot)**: points follow the reference line closely across most of the range, with only minor deviation at the tails.
- **C (residuals vs. fitted)**:  the scatter looks reasonably random, with no obvious funnel or trend. One could argue there's a faint polynomial-like curvature, but by and large the pattern appears random.
- **D (Shapiro-Wilk)**: the p-value is above 0.05, supporting the visual impression of approximate normality.

Together, these four panels indicate that our model fits the iris dataset reasonably well, with no obvious violations of the modelling assumptions. Woo-hoo!

Notice the difference from the simulated example above: real data is rarely this text-book clean, and panel C's faint curvature is a good illustration of why. `resplot()` can't tell you whether a borderline pattern like that is real structure or just noise, that call comes down to judgement, informed by things like how large the sample is, how strong the pattern looks relative to the scatter around it, and how much is riding on the conclusions drawn from the model. Here, with no funnel or clear trend accompanying it, there isn't a strong case for a more complex model, but a more cautious analyst might still investigate further (e.g. fitting a quadratic term, or checking for structure by `Species`). When a pattern is this subtle, there often isn't a single "correct" answer, only a judgement call you should be prepared to explain.

### Example: Extreme Cases

The following examples illustrate two common violations of assumption using simulated dataset so you know what to look for in your own analyses.

#### **Heteroskedasticity — variance that grows with the fitted value**

```{r het_example, fig.align='center'}
p_het <- resplot(model_het)

# Compute annotation coordinates from the model
std_resids <- rstandard(model_het)
fv <- fitted(model_het)
n <- length(std_resids)

# Panel B: red open circle around the most extreme QQ point
extreme_idx <- which.max(std_resids)
theoretical_extreme <- qnorm((rank(std_resids)[extreme_idx] - 0.5) / n)

p_het[[2]] <- p_het[[2]] +
    ggplot2::annotate(
        "point",
        x = theoretical_extreme, y = std_resids[extreme_idx],
        shape = 1, colour = "red", size = 8, stroke = 1.5
    )

# Panel C: two diverging red dashed lines indicating the funnel shape
x0 <- min(fv)
x1 <- max(fv)

p_het[[3]] <- p_het[[3]] +
    ggplot2::annotate(
        "segment",
        x = x0, xend = x1,
        y = 0.2, yend = max(std_resids) * 0.85,
        colour = "red", linewidth = 0.8, linetype = "dashed"
    ) +
    ggplot2::annotate(
        "segment",
        x = x0, xend = x1,
        y = -0.2, yend = min(std_resids) * 0.85,
        colour = "red", linewidth = 0.8, linetype = "dashed"
    )

p_het
```

Panel C shows the classic funnel shape: residual spread increases as fitted values get larger, and the Shapiro-Wilk test detects the resulting non-normality. The data here were generated as `y = mean(x) * noise`, where the noise is multiplicative, which means that the variance scales with the mean squared. 

To resolve this, alog-transformation converts this multiplicative structure into an additive one
where the variance is (approximately) constant:

```{r het_fix, fig.align='center'}
model_het_log <- lm(log(y_het) ~ x)
resplot(model_het_log)
```

After log-transforming the response, the funnel in panel C disappears: the residual scatter is now roughly uniform across fitted values. The histogram (A) and Q-Q plot (B) are both consistent with normality, and the Shapiro-Wilk p-value is comfortably above 0.05. The log scale is the appropriate scale of analysis. Fitted values and confidence intervals should therefore be back-transformed via `exp()` before results are reported on the original scale, though the back-transformed intervals require careful interpretation. In `biometryassist`, the prediction function `multiple_comparisons()` handles this automatically. See the [*Choosing and interpreting multiple comparisons*](choosing-multiple-comparisons.html) vignette for more on interpreting back-transformed intervals and choosing between comparison methods.

#### **Non-normal (right-skewed) residuals**

```{r skew_example, fig.align='center'}
resplot(model_skew)
```

Panels A and B show clear right skew: the histogram has a long right tail and th Q-Q plot follows an S-shaped curve, which are indications of model assumptions being violated. The obvious reason here is that the response was generated as `y = exp(linear + noise)`, so a log transformation again is the natural remedy it undoes the exponentiation and returns us to the linear scale on which the
data were generated:

```{r skew_fix, fig.align='center'}
model_skew_log <- lm(log(y_skew) ~ x)
resplot(model_skew_log)
```

After log-transforming, all four panels are consistent with the model assumptions: the histogram is symmetric and bell-shaped, the Q-Q plot follows the reference line closely, panel C shows random scatter with no curve or fan, and the Shapiro-Wilk test no longer rejects normality.

## Common Transformations

Analysts often settle on the right transformation during exploratory analysis, before a model is even fitted. Residual diagnostics offer a useful second check — confirming afterwards whether the transformation that was (or wasn't) applied was the appropriate one. The table below summarises common residual patterns and the transformations they typically call for:

| Residual pattern | Potential data type | Transformation | R expression |
|-----|---|---|---|
| Right-skewed histogram; S-curve in Q-Q Plot; funnel in Residual vs Fitted Plot (i.e., there is a trend in dispersion along the x axis) | Continuous positive measurements | Log | `log(y)` |
| Mild right skew; mild funnel in Residual vs Fitted Plot | Counts (Integers $\geq$ 0) | Square root | `sqrt(y)` |
| Variance highest near fitted values in the middle of the range, lowest at extremes; non-constant spread in C | Proportions / percentages | Arcsine square root | `asin(sqrt(y))` |
| Left-skewed histogram; mirror-image S-curve in Q-Q Plot; residual spread narrows as fitted values increase | Continuous measurements bounded above (e.g. scores approaching a ceiling or maximum) | Square (or reflect-and-log) | `y^2` |

The log transformation is demonstrated in detail in the [*Extreme Cases*](#example-extreme-cases) section above. The sections below walk through the other two common cases.

### Square Root Transformation (Count Data)

Count data such as numbers of insects caught, seeds germinating, or wireworms recovered follow a Poisson-like distribution in which the variance is proportional to the mean. As fitted values increase, the residual spread grows with them, which produces a mild funnel in panel C and mild right skew in panels A and B.

```{r count_raw, fig.align='center'}
set.seed(1001)
x <- runif(80, 1, 10)
y_count <- rpois(80, lambda = 1.5 * x)
m_count <- lm(y_count ~ x)

ggplot(data.frame(y_count = y_count), aes(x = y_count)) +
    geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
    labs(
        title = "Histogram of the original response, y_count",
        subtitle = "(not standardised residuals — compare with resplot() panel A below)",
        x = "Count",
        y = "Frequency"
    ) +
    theme_bw()
```

The right skew already visible in the raw counts carries through into the model's residuals. Applying `resplot()` to the fitted model gives:

```{r count_resplot, fig.align='center'}
resplot(m_count)
```

The pattern is more subtle than in the extreme cases above, but panel C still fans out gradually as fitted values increase, consistent with Poisson-type variance, while the histogram (A) is mildly right-skewed and the Shapiro-Wilk test (D) is significant. For count data that genuinely follows a Poisson distribution, Poisson regression is a natural alternative to consider instead of transforming.

Alternatively, the square root transformation can stabilise the variance directly: for a Poisson random variable $\text{Y}$ with mean $\lambda$, $\text{Var}(\sqrt{\text{Y}}) \approx \tfrac{1}{4}$ regardless of $\lambda$, so the transformed residuals should have roughly constant spread:

```{r count_sqrt, fig.align='center'}
m_count_sqrt <- lm(sqrt(y_count) ~ x)
resplot(m_count_sqrt)
```

After the square root transformation, the funnel in panel C disappears, the histogram is more symmetric, the Q-Q plot hugs the reference line more closely, and the Shapiro-Wilk p-value is no longer significant. Fitted values and confidence intervals should be back-transformed by squaring before reporting on the original count scale.

### Arcsine Square Root Transformation (Proportion Data)

Some proportion data in agriculture — for example, the percentage of barley grain retained on a sieve during grading — tend to cluster near one end of the scale, and for these a plain `log()` or `sqrt()` transformation is often perfectly reasonable. Other proportions range more freely across the full 0–1 scale, with germination rates, disease incidence, and ground cover as typical examples. The variance of a proportion $p$ estimated from $n$ trials is $p(1-p)/n$, which is largest near $p = 0.5$ and shrinks toward zero as $p$ approaches 0 or 1. This produces non-constant residual variance: observations whose true proportion sits near the middle of the range scatter widely, while those near the extremes are compressed.

```{r prop_raw, fig.align='center'}
set.seed(18)
x <- runif(80, 0, 5)
p_true <- plogis(-1 + 0.4 * x)
y_prop <- rbinom(80, size = 50, prob = p_true) / 50
m_prop <- lm(y_prop ~ x)

ggplot(data.frame(y_prop = y_prop), aes(x = y_prop)) +
    geom_histogram(bins = 15, fill = "steelblue", colour = "white") +
    labs(
        title = "Histogram of the original response, y_prop",
        subtitle = "(not standardised residuals — compare with resplot() panel A below)",
        x = "Proportion",
        y = "Frequency"
    ) +
    theme_bw()
```

Applying `resplot()` to the fitted model without transforming the response variable gives:

```{r prop_resplot, fig.align='center'}
resplot(m_prop)
```

Panel C shows non-constant spread — residuals are more variable at intermediate fitted values and narrow at the high and low ends. The arcsine square root ("angular") transformation $y' = \arcsin(\sqrt{p})$ is designed to stabilise this variance:

```{r prop_asin, fig.align='center'}
m_prop_asin <- lm(asin(sqrt(y_prop)) ~ x)
resplot(m_prop_asin)
```

After the transformation, the spread in panel C is more uniform across fitted values and the histogram is closer to symmetric — though the improvement is more subtle here than in the count-data example above, and reasonable analysts could disagree on whether it's enough to be worth it. This is where subjective judgement comes in: there's no fixed threshold at which a residual pattern definitively "passes", so weighing this modest gain in diagnostics against the cost of interpreting results on the arcsine-square-root scale is ultimately a judgement call.

The arcsine square root is most effective when proportions are not too extreme (roughly 0.2–0.8) and when each observation is based on a reasonable number of trials. When proportions cluster near 0 or 1, or when the underlying counts are very small, a logistic mixed model is usually more appropriate than a variance-stabilising transformation.

## Simple Field Trial Example

To illustrate how `resplot()` fits into a real analysis workflow, we use `cochran.wireworms` from the `agridat` package, a classic soil fumigation experiment from @cochran1938. Five fumigant treatments (K, M, N, O, and an untreated control P) were applied to 5 × 5 soil sections arranged in a Latin Square: each treatment appears exactly once in every row and every column, so both spatial dimensions serve as blocking factors.

```{r real_data_setup, message=FALSE}
library(agridat)
data(cochran.wireworms)
head(cochran.wireworms)
```

The response variable `worms` records the number of wireworms recovered from
each soil section after the fumigation period.

### A naïve model and its residuals

As a starting point, suppose we fit a simple one-way model that includes only the treatment factor, ignoring the Latin Square structure:

```{r real_data_naive}
m_naive <- aov(worms ~ trt, data = cochran.wireworms)
resplot(m_naive)
```

The diagnostic panels raise an immediate concern. The histogram (A) is visibly right-skewed, the Q-Q plot (B) shows the upper tail pulling away from the reference line, and the Shapiro-Wilk test (D) returns a p-value well below 0.05 — all pointing to non-normal residuals.

Before reaching for a transformation, it is worth asking *why* the residuals look this way. A quick tabulation of mean worm counts by row reveals substantial row-to-row variation:

```{r real_data_row_means}
tapply(cochran.wireworms$worms, cochran.wireworms$row, mean)
```

Rows 1–3 average fewer than 4 worms per section, while row 4 averages well above that. This is exactly the kind of spatial gradient that the Latin Square row-blocking is designed to absorb — but the naïve model has left it entirely in the residuals, inflating a handful of them and distorting the residual
distribution.

### Accounting for the Latin Square Design

Including `row` and `col` as blocking factors gives the model the Latin Square intended to provide:

```{r real_data_ls, fig.align='center'}
m_ls <- aov(worms ~ trt + factor(row) + factor(col), data = cochran.wireworms)
resplot(m_ls)
```

The picture changes substantially:

- **A (histogram)**: now approximately symmetric and bell-shaped.
- **B (Q-Q plot)**: points follow the reference line with only minor tail deviation — typical for $n = 12$ residuals after removing 12 degrees of freedom for blocking.
- **C (residuals vs. fitted)**: scatter is random with no obvious structure.
- **D (Shapiro-Wilk)**: p-value is now well above 0.05, consistent with normally distributed residuals.

The residual diagnostics did not point to a distributional problem requiring transformation; they flagged a misspecified model. Accounting for the blocking structure in the design resolved the issue and, as a practical bonus, also increases precision for the treatment comparisons by removing the spatial nuisance variation from the error.

## Chickpea Nitrogen Fixation Trial Example (ASReml-R)

We will be using a real data example which involves a more complex setting. The `dat` extracted below contains the dataset for a split-plot greenhouse pot trial investigating rhizobium strains for biological nitrogen fixation in chickpea (*Cicer arietinum*). The trial compared 13 rhizobium strains across 8 chickpea cultivars, giving 416 observations of shoot dry weight (mg/plant) arranged in replicate blocks along rows and columns of a greenhouse bench. 

The code in this section requires **ASReml-R**, a commercial mixed model package. To run this code yourself, you may need a valid ASReml-R licence from [VSNi](https://vsni.co.uk/). We start by doing some data processing before fitting:

```{r chickpea_data, message=FALSE}
library(openxlsx2)
library(biometryassist)
biometryassist:::quiet(library(asreml))


# Importing Data
dat <- read_xlsx(
    "resplot-example/Latte pot method chickpea x Mciceri test.xlsx",
    sheet = "Stats Worksheet",
    na.strings = "*"
)

# Standardise column names to snake_case
names(dat) <- tolower(gsub("[ /]", "_", names(dat)))

# Create integer row/col indices for the spatial correlation structure.
# Rows run within each rep/block/table; cols are encoded in the pot number.
dat$row <- factor(((as.integer(dat$rep_block_table) - 1L) * 8L) + as.integer(dat$pot_row))
dat$col <- factor(as.integer(substr(dat$pot_number, 2, 3)))
dat <- dat[order(dat$row, dat$col), ]

# Remove zero shoot dry weight (structural zeros, not meaningful measurements)
dat$shoot_dw_mg_plant[dat$shoot_dw_mg_plant == 0] <- NA

# Ensure treatment factors are correctly typed
dat$strain   <- factor(dat$strain)
dat$cultivar <- factor(dat$cultivar)
dat$rep_block_table <- factor(dat$rep_block_table)
dat$mainplot <- factor(dat$mainplot)
```

The trial was laid out on 4 greenhouse benches, each holding one complete replicate of the design: `rep_block_table` records which of the 4 benches ("tables") a pot sits on, `mainplot` is the whole-plot within that bench, with one per rhizobium strain being the main plot. `pot_row` gives the position of each of the 8 cultivar sub-plots within a mainplot. Together, `rep_block_table` and `pot_row` build the `row` index above, while `col` (from the last two digits of `pot_number`) records which mainplot/strain column a pot belongs to. The raw spreadsheet also carries a few columns not needed for this analysis (a row `id`, nodulation counts, %SP), so before continuing we keep just what's used below:

```{r chickpea_clean}
dat <- dat[, c(
    "rep_block_table", "mainplot", "row", "col",
    "strain", "cultivar", "shoot_dw_mg_plant"
)]
```

Sometimes it is also good practice to visualise the experimental design element to see how things are laid out, if it is not immediately obvious. Plotting each pot at its physical `row`/`col` position on the benches, coloured by `strain` (the whole-plot factor) and labelled by `cultivar` (the sub-plot factor, first letter of its name), shows the split-plot structure directly:

```{r chickpea_design_plot, fig.width=7, fig.height=6, fig.align='center'}
ggplot(dat, aes(x = col, y = row, fill = strain)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = substr(cultivar, 1, 1)), size = 2) +
    geom_hline(yintercept = c(8.5, 16.5, 24.5), linewidth = 0.8) +
    scale_y_discrete(limits = rev) +
    scale_fill_discrete() +
    labs(
        x = "Column (mainplot)", y = "Row",
        caption = "Layout of the 4 greenhouse benches. Tile colour = strain (whole-plot); letter = cultivar (sub-plot). Horizontal lines mark the 4 rep_block_table benches"
    ) +
    theme_bw()
```

Each horizontal band between the reference lines is one bench (`rep_block_table`), and each column within a band is one mainplot: all 8 tiles in that column share the same strain colour but carry 8 different cultivar letters (C = Captain, D = Drummond, G = Gen090, H = Hattrick, K = Kyabra, M = Magnus, R = Royal, S = Seamer), since cultivar is randomised as the sub-plot factor within each strain's mainplot. The strain colours scramble from one band to the next, confirming that strains were randomised to mainplots independently within each bench — this is a randomised complete block, split-plot design: every bench is a complete replicate containing all 13 strains fully crossed with all 8 cultivars, not confounded with each other, just estimated at different strata (whole-plot for strain, sub-plot for cultivar).

### Exploratory Data Analysis

While we have largely omitted the EDA step for the previous simple examples, in real life it is often good to observe what the data look like before fitting the model. This is especially the case for experimental design data, where the design structure may not be immediately obvious. Nonetheless, we first investigate the shape of our response variable `shoot_dw_mg_plant`, which records the shoot dry matter measurement per plant.

```{r}
head(dat)
```

```{r chickpea_hist, fig.width=7, fig.height=3, fig.align='center'}
ggplot(dat, aes(x = shoot_dw_mg_plant)) +
    geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
    labs(x = "Shoot dry weight (mg/plant)", y = "Frequency") +
    theme_bw()
```

Comparing the distribution of shoot dry weight across strains and cultivars gives an early read on which treatment factors are likely to matter. Rather than building these by hand, `biometryassist` already ships `summary_graph()` for exactly this kind of comparison:

```{r chickpea_summary_strain, fig.width=7, fig.height=4, fig.align='center'}
summary_graph(dat, "shoot_dw_mg_plant", "strain", "mg/plant") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r chickpea_summary_cultivar, fig.width=7, fig.height=4, fig.align='center'}
summary_graph(dat, "shoot_dw_mg_plant", "cultivar", "mg/plant")
```

Both factors look like they shift the mean shoot dry weight, which is the basic case for including `strain` and `cultivar` as fixed effects at all. That still leaves the question of whether they act independently or interact, , which is what justifies fitting `strain * cultivar` (with an interaction term) rather than just `strain + cultivar` (main effects only) in every model below. Passing two variables to `summary_graph()` instead of one switches it to an interaction-style plot: the mean of each level of the second variable is traced as a line across the levels of the first, so a genuine interaction shows up as lines that cross or diverge, while lines that stay roughly parallel indicate the two factors act additively.

```{r chickpea_summary_interaction, fig.width=7, fig.height=4, fig.align='center'}
summary_graph(dat, "shoot_dw_mg_plant", c("strain", "cultivar"), "mg/plant") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r chickpea_summary_interaction2, fig.width=7, fig.height=4, fig.align='center'}
summary_graph(dat, "shoot_dw_mg_plant", c("cultivar", "strain"), "mg/plant")
```

### Model Fitting Process

Having done the exploratory analysis, we move into modelling with a clearer picture: the response likely needs a `log()` or `sqrt()` transformation, and strain/cultivar may need some heterogeneous variance treatment. But suppose we hadn't looked at any of that yet — a naive first attempt would model just the two treatment effects and their interaction:

```{r model_simple, warning=FALSE, fig.align='center'}
model_simple <- asreml(
    shoot_dw_mg_plant ~ strain * cultivar,
    data = dat,
    trace = FALSE
)

resplot(model_simple)
```

This clearly presents some problems. Before picking a transformation, it helps to compare the response on a few candidate scales side by side:

```{r chickpea_transform_compare, warning=FALSE, fig.width=9, fig.height=3, fig.align='center'}
p_raw <- ggplot(dat, aes(x = shoot_dw_mg_plant)) +
    geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
    labs(title = "Original scale", x = "Shoot dry weight (mg/plant)") +
    theme_bw()

p_log <- ggplot(dat, aes(x = log(shoot_dw_mg_plant))) +
    geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
    labs(title = "Log scale", x = "log(shoot dry weight)") +
    theme_bw()

p_sqrt <- ggplot(dat, aes(x = sqrt(shoot_dw_mg_plant))) +
    geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
    labs(title = "Square root scale", x = "sqrt(shoot dry weight)") +
    theme_bw()

p_raw + p_log + p_sqrt
```

Of the three, the log-transformed histogram is the most symmetric and bell-shaped, so we proceed with the log transformation for the remaining models. Refitting with `log(shoot_dw_mg_plant)` as the response, the dispersion in the residuals-vs-fitted panel is already noticeably less pronounced:

```{r model_2, warning=FALSE, fig.align='center'}
model_2 <- asreml(
   log(shoot_dw_mg_plant) ~ strain * cultivar,
    data = dat,
    trace = FALSE
)

resplot(model_2)
```

The resplot above still needs some work. `model_2` treats every plot as independent, but the trial actually has an experimental design behind it — a randomised complete block, split-plot structure, with `rep_block_table` as the block and `mainplot` as the whole-plot. Fitting these as random effects lets each block and each mainplot have its own deviation from the overall mean, rather than forcing all of that block-to-block and mainplot-to-mainplot variation into the residuals, so accounting for them should tighten things up further:

```{r model_3, warning=FALSE, fig.align='center'}
model_3 <- asreml(
   log(shoot_dw_mg_plant) ~ strain * cultivar,
   random = ~ rep_block_table / mainplot,
    data = dat,
    trace = FALSE
)

resplot(model_3)
```

The central cluster in panel C now looks more random than before. However, the Q-Q plot still shows heavy tails at both ends. Before reaching for per-group variance terms on faith, it's worth checking whether specific strains or cultivars are actually driving that tail-heaviness: plotting `model_3`'s residuals grouped by strain and by cultivar shows whether some groups are noticeably more (or less) variable than the rest.

```{r model_3_resid_by_group, warning=FALSE, fig.width=7, fig.height=7, fig.align='center'}
resid_dat <- dat
resid_dat$resid <- residuals(model_3)

het_strains   <- c("MC4", "CC1192", "AU18(4)", "MI_8", "WSM5041")
het_cultivars <- c("Royal", "Magnus")

p_resid_strain <- ggplot(resid_dat, aes(x = strain, y = resid, fill = strain %in% het_strains)) +
    geom_violin(alpha = 0.6, colour = NA, show.legend = FALSE) +
    geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
    scale_fill_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey80")) +
    labs(x = "Strain", y = "model_3 residual") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p_resid_cultivar <- ggplot(resid_dat, aes(x = cultivar, y = resid, fill = cultivar %in% het_cultivars)) +
    geom_violin(alpha = 0.6, colour = NA, show.legend = FALSE) +
    geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
    scale_fill_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey80")) +
    labs(x = "Cultivar", y = "model_3 residual") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p_resid_strain / p_resid_cultivar
```

The boxes highlighted in red are the strains and cultivars given their own residual variance in `model_4` below (`at(strain, "MC4"):units`, and so on) — check whether they do in fact show visibly wider or narrower spread than the grey boxes. That comparison, on the model's own (log) scale and after the fixed and block effects are already accounted for, is the actual evidence for adding `at():units` terms, rather than the raw pre-model boxplots from the EDA section above (which are on the untransformed scale and reflect total spread, not residual spread).

```{r model_4, warning=FALSE}
model_4 <- asreml(
    log(shoot_dw_mg_plant) ~ strain * cultivar,
    random = ~ rep_block_table / mainplot +
        at(strain, "MC4"):units +
        at(strain, "CC1192"):units +
        at(strain, "AU18(4)"):units +
        at(strain, "MI_8"):units +
        at(strain, "WSM5041"):units +
        at(cultivar, "Royal"):units +
        at(cultivar, "Magnus"):units,
    data = dat,
    trace = FALSE
)

resplot(model_4)
```

The `at(factor, "level"):units` terms let selected strains and cultivars have their own residual variance, rather than assuming every group is equally variable — the diagnostic above is what actually flags `CC1192` and `MC4` (among strains) as having noticeably different spread from the rest. `Magnus` is included for a different reason: it had several missing values, which changes its effective precision regardless of whether its residual spread stands out. Allowing for this heterogeneity, rather than assuming every group shares one common residual variance, is what tightens the Q-Q tails compared to `model_3`.

Although not covered in detail elsewhere in this vignette, the physical location of a plant can also matter: pots that sit next to each other on the bench tend to share a similar micro-environment, a phenomenon known as spatial autocorrelation. In ASReml-R, this can be added directly with `residual = ar1(row):ar1(col)`.

```{r chickpea_model, warning=FALSE}
model.asr <- asreml(
    log(shoot_dw_mg_plant) ~ strain * cultivar,
    random = ~ rep_block_table / mainplot +
        at(strain, "MC4"):units +
        at(strain, "CC1192"):units +
        at(strain, "AU18(4)"):units +
        at(strain, "MI_8"):units +
        at(strain, "WSM5041"):units +
        at(cultivar, "Royal"):units +
        at(cultivar, "Magnus"):units,
    residual = ~ ar1(row):ar1(col),
    data = dat,
    trace = FALSE
)
```

```{r chickpea_resplot, fig.align='center'}
resplot(model.asr)
```

Overall, the final resplot aligns well with our modelling assumptions — dare we say, beautiful. The Shapiro-Wilk test still flags the residuals as non-normal, but with a relatively large sample (n = 406), the test has high power: minor, practically unimportant departures from normality get flagged as significant just as readily as meaningful ones, so this result can be taken with a grain of salt rather than a cause for concern.

For a well-fitting model on these data, we would expect the histogram (panel A) to be approximately symmetric and bell-shaped around zero, the Q-Q plot (panel B) to sit close to the reference line with only modest deviation in the tails, and the residuals-vs-fitted plot (panel C) to show a horizontal scatter with no funnel shape or systematic curve — confirming that the `ar1` spatial structure and the per-group variance components have absorbed the main sources of heterogeneity in the data.

Modelling is an iterative process: we try to incorporate relevant information into the analysis, and while any single addition might not help much on its own, together they can amount to a substantial improvement, as shown here.

## Playing Around with the Visuals

`resplot()`'s output follows normal `ggplot2` styling conventions, so it can be customised and extended using familiar syntax. But because it returns a `patchwork` object built from ggplot2 panels — not a single `ggplot` object — that customisation happens through patchwork's own syntax:

- Use **`&`** to apply a ggplot2 theme or element to *every* panel at once.
- **Index** into an individual panel (e.g. `resplot(model)[[1]]`) to modify it with normal ggplot2 `+` syntax.

Patchwork reserves `+` for adding a *new* panel to the layout, so `resplot(model) + theme_bw()` won't restyle the existing plots the way it would for a single `ggplot` object — reach for `&` instead.

### Restyle every panel at once with `&`

```{r theme_all_panels, fig.align='center'}
resplot(model, shapiro = FALSE) & theme_classic()
```

One line swaps `resplot()`'s default `theme_bw()` for `theme_classic()` across all three panels simultaneously, handy for matching diagnostics to the theme of a report or slide deck.

### Changing the data mark colours

The fill and outline colours of the histogram bars and scatter points are hard-coded inside `resplot()`, so neither ggplot2 scales nor `update_geom_defaults()` can override them — both are bypassed by the explicit static aesthetics the function sets on each layer. The way around this is to reach into each layer directly after the plot has been built, and overwrite those static aesthetics yourself.

Each panel returned by `resplot()` is a standard `ggplot` object whose first layer holds the histogram bars or scatter points; the static colours live in `$layers[[1]]$aes_params`. Overwrite those entries before printing the panel:

```{r recolour_one_panel, fig.align='center'}
p <- resplot(model, shapiro = FALSE)[[1]]  # histogram panel (A)
p$layers[[1]]$aes_params$fill <- "steelblue"
p$layers[[1]]$aes_params$colour <- "navy"
p
```

To recolour all three panels at once, apply the same replacement to each one inside a loop — patchwork's indexing lets you modify a panel of an already-composed object in place, so no reassembly is needed:

```{r recolour_all_panels, message=FALSE, fig.align='center'}
recolour_resplot <- function(plt, fill = "steelblue", colour = "navy") {
    for (i in 1:3) {
        plt[[i]]$layers[[1]]$aes_params$fill <- fill
        plt[[i]]$layers[[1]]$aes_params$colour <- colour
    }
    plt
}

recolour_resplot(resplot(model, shapiro = FALSE))
```

The helper loops over the three panels (`[[1]]` histogram, `[[2]]` Q-Q plot, `[[3]]` residuals-vs-fitted), overwrites the stored static aesthetics on each, and returns the same patchwork object with the new colours applied. Combine it freely with `&` for theme changes or `/` for model comparisons.

### Annotate a single panel by indexing

```{r annotate_one_panel, message=FALSE, fig.align='center'}
resplot(model, shapiro = FALSE)[[3]] +
    geom_smooth(se = FALSE, colour = "firebrick", linewidth = 0.5) +
    labs(title = "Residuals vs. Fitted, with a smoother")
```

Once extracted, a panel is an ordinary `ggplot` object, so any ggplot2 layer, scale or theme element can be added directly with `+`. Here, overlaying a loess smoother on panel C makes a subtle curved trend easier to spot than the raw scatter alone.

### Compare two models' diagnostics in one figure

```{r compare_two_models, fig.width=9, fig.height=8, fig.align='center'}
model_a <- aov(Petal.Length ~ Petal.Width, data = iris)
model_b <- aov(Petal.Length ~ Petal.Width + Species, data = iris)

(resplot(model_a, shapiro = FALSE) & theme_minimal()) /
    (resplot(model_b, shapiro = FALSE) & theme_minimal())
```

Because each `resplot()` call already returns a patchwork object, two (or more) of them can be composed further with patchwork's own `/` (stack) and `|` (side-by-side) operators — useful here for eyeballing whether adding `Species` to the model actually improved the residual behaviour.

### Showing the Model Call

Set `call = TRUE` to print the model's call as a title above the plots, handy when comparing diagnostics from several candidate models side by side. Use `call.size` to adjust its font size:

```{r call_true, fig.align='center'}
resplot(model, call = TRUE, call.size = 8)
```

### Adjusting Text Sizes

Two further size arguments control the rest of the plot's typography:

- `label.size` — the font size of the panel tags (`A`, `B`, `C`).
- `axes.size` — the font size of the axis text and titles within each panel.

```{r sizing, fig.align='center'}
resplot(model, label.size = 14, axes.size = 8)
```

### Multi-Stratum Models

Some designs are analysed with more than one error stratum. For example, a split-plot design fitted with `aov()` and an `Error()` term, or an ASReml-R model with multiple residual structures. For these, a single histogram/Q-Q/fitted-values panel can't summarise every stratum, so `resplot()` returns a
**list** of plots, one per stratum, named after the strata.

```{r split_plot_setup}
set.seed(44)
n_blocks <- 8
split_plot_dat <- data.frame(
  block = factor(rep(1:n_blocks, each = 6)),
  irrigation = factor(rep(
    rep(c("Irrigated", "Rainfed"), each = 3),
    times = n_blocks
  )),
  variety = factor(rep(c("A", "B", "C"), times = 2 * n_blocks))
)
split_plot_dat$yield <- 50 +
  5 * (split_plot_dat$irrigation == "Irrigated") +
  c(A = 0, B = 2, C = 4)[split_plot_dat$variety] +
  rnorm(nrow(split_plot_dat), sd = 3)

split_plot_model <- aov(
  yield ~ irrigation * variety + Error(block / irrigation),
  data = split_plot_dat
)
```

```{r split_plot_resplot, fig.align='center'}
split_plot_resids <- resplot(split_plot_model, shapiro = FALSE)
names(split_plot_resids)
```

Each element is a complete diagnostic panel for that stratum, and can be
inspected (or saved) individually:

```{r split_plot_print, fig.align='center'}
split_plot_resids[["block:irrigation"]]
```

The same applies to multi-stratum ASReml-R models — each residual variance
structure becomes its own named element in the returned list.

### Putting Multiple Strata on One Page

When a model has several strata, printing each one separately is tedious. Set `onepage = TRUE` to arrange them into a grid (up to 6 strata per page, paginated automatically if there are more), and use `onepage_cols` to control the number of columns. With three strata, `onepage_cols = 1` stacks them in a single column:

```{r onepage_example, fig.width=7, fig.height=15, fig.align='center'}
resplot(
  split_plot_model,
  shapiro = FALSE,
  onepage = TRUE,
  onepage_cols = 1
)
```

`onepage_cols` matters more once there are enough strata to need more than one row — for a model with, say, 5 strata, `onepage_cols = 2` would arrange them as two rows of two plus a final row of one, rather than `resplot()`'s default 3-column grid. If there are more than 6 strata in total, `resplot()` returns a *list of pages* instead of a single combined plot — index into it (`pages[[1]]`, `pages[[2]]`, ...) to view each page.

## Supported Model Types {#supported-model-types}

`resplot()` produces diagnostics for any model class with an `extract_model_info()` method. At the time of writing these are: `aov`, `lm`, `aovlist` (multi-stratum `aov`), `lme` ([nlme::lme()]), `lmerMod`/
`lmerModLmerTest` ([lme4::lmer()], `lmerTest::lmer()`, and relationship-based `lme4breeding::lmebreed()` models), `asreml`, `mmer`/`mmes` (sommer), `art` (ARTool), `afex_aov` (afex) and `glmmTMB` (Gaussian family only — non-Gaussian families should use `DHARMa::simulateResiduals()` instead, since a normal Q-Q
plot isn't a valid diagnostic for them).

This is a slightly broader set than the comparison functions documented in the [*Choosing and interpreting multiple comparisons*](choosing-multiple-comparisons.html) vignette: `resplot()` additionally supports ARTool models and sommer's legacy `mmer` interface, neither of which has a `predict()`-based comparison equivalent.

See `?resplot` for the full, authoritative table, and `?extract_model_info` if
you need to add support for another model engine.


## Troubleshooting [@TODO]

**"`model.obj` must be a linear (mixed) model object..."** — the object you
passed isn't one of the supported classes listed above. Check `class(your_model)`
against the supported list; unsupported engines need a new
`extract_model_info.<class>()` method before `resplot()` can handle them.

**The Shapiro-Wilk panel is missing** — either you set `shapiro = FALSE`, or
your model has 5,000 or more residuals, in which case the test is always
skipped (see [*D — Shapiro-Wilk Test*](#d-shapiro-wilk-test)).

**Warnings about non-finite or missing values being removed** — this usually
means some residuals or fitted values are `NA` (for example, from a
rank-deficient fit or missing data left in the model frame). Inspect
`residuals(your_model)` and `fitted(your_model)` directly to find the cause
before trusting the diagnostics.

**Plots look cramped with `onepage = TRUE`** — reduce `onepage_cols`, or omit
`onepage` and review the strata one at a time instead.


## Further Reading [@TODO]

- `?resplot` for the complete argument and supported-model reference.
- `?extract_model_info` for the extension mechanism used to add new model engines.
- The [*Choosing and interpreting multiple comparisons*](choosing-multiple-comparisons.html)
  vignette, for the next step after the model passes these checks.

## References

::: {#refs}
:::
