---
title: "Diagnosing Model Fit with resplot()"
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-depth: 3
    toc-location: right   # or "right", "body"
    code-fold: true       # collapsible code blocks
    code-tools: true      # copy button + show/hide all
    number-sections: false
    fig-width: 7
    fig-height: 5
    embed-resources: true 
vignette: >
  %\VignetteIndexEntry{Diagnosing Model Fit with resplot()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
engine: knitr
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, message=FALSE}
library(biometryassist)
library(ggplot2)
library(patchwork)
```

## Introduction

After fitting a model to data, it is important to check the residuals — the leftover variation in the data to gain important insight. In particular, we want to check whether the fitted model is appropriate for the data and adheres to key modelling assumptions. These assumptions may include approximately normal and homogeneous residual variance, and linearity in the case of linear regression. In `biometryassist`, `resplot()` is a function that provides a single-panel diagnostic plot for a fitted model.

A call to `resplot()` always returns four core panels, built from the model's
*standardised* residuals (raw residuals divided by their standard deviation):

- **A — Histogram** of the standardised residuals, to check the overall shape of
  the residual distribution.
- **B — Normal Q-Q plot**, to check whether the residuals follow a normal
  distribution, especially in the tails.
- **C — Residuals vs. fitted values**, to check for non-constant variance or
  remaining structure that the model hasn't captured.
- **D — A Shapiro-Wilk normality test**, displayed as text with relevant p-values and conclusion underneath the three plots (toggle with and `shapiro = TRUE` as the default).

```{r basic_example}
model <- aov(Petal.Length ~ Petal.Width, data = iris)
resplot(model)
```
<!-- Need Check -->
`resplot()` works across most of the modelling engines supported by
`biometryassist` — `aov`/`lm`, `nlme::lme()`, `lme4`/`lmerTest`, `ASReml-R`,
`sommer`, `ARTool` and others, by dispatching on the model's class via an internal
`extract_model_info()` generic. See the
[*Supported model types*](#supported-model-types) section below.

## Understanding the Panels

Each call to `resplot()` returns three core diagnostic panels (A–C) assembled from
the model's standardised residuals, plus an optional Shapiro-Wilk result (D).
Together they give a consistent view of how well a model satisfies the normality
and constant-variance assumptions. Each panel below is pulled directly from
`resplot()`'s output for the example models featured in the
[*Full Output Example*](#full-output-example) section below using `resplot(model)[[i]]`, with `i = 1, 2, 3` corresponding to the three plots.

```{r example_models_setup}
set.seed(42)
n <- 100
x <- runif(n, 1, 10)
y_het <- 2 + 0.5 * x + rnorm(n, sd = x * 0.3)
model_het <- lm(y_het ~ x)

set.seed(99)
x <- runif(100, 1, 10)
y_skew <- exp(1 + 0.3 * x + rnorm(100, sd = 0.5))
model_skew <- lm(y_skew ~ x)
```

Alongside the well-behaved `model` fitted in the Introduction, `model_het` and
`model_skew` are built so that each breaks the model's assumptions in a
different, deliberate way: `model_het`'s errors are simulated with
`sd = x * 0.3`, so their spread grows with `x`, while `model_skew`'s errors are
exponentiated, giving them a right-skewed distribution instead of a normal one.

A linear (mixed) model assumes

$$ y = X\beta + Zu + e, \qquad e \sim N(0, \sigma^2 I) $$

that is, the errors $e$ are independent, normally distributed, and share a
single constant variance $\sigma^2$. `model_het`'s growing error spread
violates the constant-$\sigma^2$ half of this assumption, and `model_skew`'s
skewed errors violate the normal half. This isn't just a technicality:
the standard errors, confidence intervals, and t-/F-tests used to compare
treatment means are all derived assuming the assumption holds, so for models
like these, the p-values and confidence intervals `resplot()` and downstream
comparisons report can end up too liberal or too conservative to trust.

We can't observe $e$ directly, so we check the assumption using the model's
standardised residuals,

$$ r_i^{\text{std}} = \frac{r_i}{\hat{\sigma}\sqrt{(n-1)/n}} $$

where $r_i$ is the raw residual for observation $i$ and $\hat\sigma$ is their
sample standard deviation. If the model's assumptions hold, $r_i^{\text{std}}$
should behave like an independent random sample from $N(0,1)$, which
`model`'s residuals broadly do, unlike `model_het` or `model_skew` by design.
Panels A–D below each check a different aspect of that one claim.

### A — Histogram

The histogram shows the distribution of the standardised residuals $r_i^{\text{std}}$.
Use it to judge the **overall shape** of the residual distribution against the
bell-shaped $N(0,1)$ density it should follow.

```{r panel_a_example, fig.width=7, fig.height=3}
p_hist_good <- resplot(model, shapiro = FALSE)[[1]] +
    labs(title = "Good: symmetric, bell-shaped")

p_hist_bad <- resplot(model_skew, shapiro = FALSE)[[1]] +
    labs(title = "Warning: right-skewed")

p_hist_good + p_hist_bad
```

A roughly symmetric, bell-shaped histogram centred on zero (left) is consistent
with normally distributed residuals. Strong right or left skew, a bimodal
distribution, or very heavy tails (right) suggest the model is missing something —
a transformation of the response, an omitted covariate, or an unaccounted-for
grouping factor.

### B — Normal Q-Q Plot

The Q-Q plot compares the empirical quantiles of $r_i^{\text{std}}$ to the
theoretical quantiles of $N(0,1)$ — the most direct check of the normality half
of the assumption. If the residuals are normally distributed, the points should
fall close to the diagonal reference line.

```{r panel_b_example, fig.width=7, fig.height=3}
p_qq_good <- resplot(model, shapiro = FALSE)[[2]] +
    labs(title = "Good: points on the reference line")

p_qq_bad <- resplot(model_skew, shapiro = FALSE)[[2]] +
    labs(title = "Warning: S-curve (skew)")

p_qq_good + p_qq_bad
```

When residuals are well-behaved (left), the points hug the diagonal line. Common
warning patterns include:

- **S-shaped curve** (right): the residuals are skewed — one tail is longer than
  the other.
- **Both ends curling away from the line**: heavy tails — more extreme values than
  a normal distribution would produce.
- **A single isolated point far from the rest**: a potential outlier that may be
  distorting the fit.

### C — Residuals vs. Fitted Values

This scatterplot places $r_i^{\text{std}}$ on the vertical axis and the model's
fitted values on the horizontal axis. It checks the constant-variance half of
the assumption: whether $\text{Var}(r_i^{\text{std}})$ stays flat across fitted
values, as $\sigma^2$ in $e \sim N(0, \sigma^2 I)$ is assumed constant rather
than a function of $\hat{y}_i$.

```{r panel_c_example, fig.width=7, fig.height=3}
p_rf_good <- resplot(model, shapiro = FALSE)[[3]] +
    labs(title = "Good: random scatter around zero")

p_rf_bad <- resplot(model_het, shapiro = FALSE)[[3]] +
    labs(title = "Warning: funnel (heteroskedasticity)")

p_rf_good + p_rf_bad
```

Ideally the points form a random cloud centred around zero with roughly constant
spread across all fitted values (left). Common problems to watch for:

- **Funnel or fan shape** (right): residual variance increases (or decreases) with
  the fitted value — heteroskedasticity. A variance-stabilising transformation
  (e.g. log, square-root) or a weighted model may be needed.
- **Curved trend**: a polynomial or non-linear term is missing from the model.
- **Distinct bands or clusters**: a factor that should be in the model has been
  omitted or mislabelled.

### D — Shapiro-Wilk Test

Below the three plots, `resplot()` (by default) displays the result of a
Shapiro-Wilk normality test on the raw residuals — a formal test of
$H_0: r_i^{\text{std}} \sim N(0,1)$, complementing the visual evidence in
panels A and B. The output always includes the test statistic and a
plain-language interpretation.

When the test finds no evidence of non-normality (p > 0.05):

```
Shapiro-Wilk normality test p-value: 0.3241
The residuals appear to be normally distributed. (n = 48)
```

When the test detects significant non-normality (p ≤ 0.05):

```
Shapiro-Wilk normality test p-value: 0.0003
The residuals do not appear to be normally distributed. (n = 48)
```

Set `shapiro = FALSE` to omit it entirely:

```{r shapiro_off}
resplot(model, shapiro = FALSE)
```

The Shapiro-Wilk test becomes unreliable as a guide for larger samples — with
enough observations, even trivial, practically meaningless departures from
normality become "significant". `resplot()` reflects this:

- For 2,000–4,999 observations, the test still runs but a warning is issued.
- For 5,000 or more observations, the test is skipped (with a warning) and
  panel D is omitted regardless of the `shapiro` argument.

In both cases, prefer to judge normality visually from panels A and B rather
than relying on the test alone.

### Full Output Example

#### Relatively good residuals

Fitting a simple one-way model to the built-in `iris` data gives residuals that are
broadly consistent with the model assumptions — a useful baseline:

```{r basic_example_2}
resplot(model)
```

Reading each panel in this output:

- **A (histogram)**: roughly symmetric and bell-shaped — consistent with normally
  distributed residuals.
- **B (Q-Q plot)**: points follow the reference line closely across most of the
  range, with only minor deviation at the tails.
- **C (residuals vs. fitted)**: the scatter is reasonably random with no obvious
  funnel or trend.
- **D (Shapiro-Wilk)**: the p-value is above 0.05, supporting the visual
  impression of approximate normality.

#### Potential extreme cases

The following examples illustrate two common violations so you know what to look
for in your own analyses.

**Heteroskedasticity — variance that grows with the fitted value**

```{r het_example}
resplot(model_het)
```

Panel C shows the classic funnel shape: residual spread increases as fitted values
get larger. Possible remedies include a log or square-root transformation of the
response, or fitting a weighted linear model.

**Non-normal (right-skewed) residuals**

```{r skew_example}
resplot(model_skew)
```

Panels A and B show clear right skew: the histogram has a long right tail and the
Q-Q plot follows an S-shaped curve. Log-transforming the response before fitting
(`log(y_skew)`) is typically the first step to investigate.

## Real Data Example


## Playing Around with the Visuals

The default styling `resplot()` ships with is only a starting point, not a
ceiling. `resplot()` returns a `patchwork` object built from ggplot2 panels, so
it can be further customised using patchwork syntax:

- Use **`&`** to apply a ggplot2 theme or element to *every* panel at once.
- **Index** into an individual panel (e.g. `resplot(model)[[1]]`) to modify it
  with normal ggplot2 `+` syntax.

Patchwork reserves `+` for adding a *new* panel to the layout, so
`resplot(model) + theme_bw()` won't restyle the existing plots the way it would
for a single `ggplot` object — reach for `&` instead.

**Restyle every panel at once with `&`**

```{r theme_all_panels}
resplot(model, shapiro = FALSE) & theme_minimal()
```

One line swaps `resplot()`'s default `theme_bw()` for `theme_minimal()` across
the histogram, Q-Q plot and residuals-vs-fitted panel simultaneously — handy for
matching diagnostics to the theme of a report or slide deck.

**Recolour or annotate a single panel by indexing**

```{r recolour_one_panel, message=FALSE}
resplot(model, shapiro = FALSE)[[3]] +
    geom_smooth(se = FALSE, colour = "firebrick", linewidth = 0.5) +
    labs(title = "Residuals vs. Fitted, with a smoother")
```

Panels are indexed in the order `resplot()` builds them — `[[1]]` is the
histogram (A), `[[2]]` the Q-Q plot (B), and `[[3]]` the residuals-vs-fitted
scatter (C). Once extracted, the panel is an ordinary `ggplot` object, so any
ggplot2 layer, scale or theme element can be added directly. Here, overlaying a
loess smoother on panel C makes a subtle curved trend easier to spot than the
raw scatter alone.

**Compare two models' diagnostics in one figure**

```{r compare_two_models, fig.width=9, fig.height=8}
model_a <- aov(Petal.Length ~ Petal.Width, data = iris)
model_b <- aov(Petal.Length ~ Petal.Width + Species, data = iris)

(resplot(model_a, shapiro = FALSE) & theme_minimal()) /
    (resplot(model_b, shapiro = FALSE) & theme_minimal())
```

Because each `resplot()` output is itself a patchwork object, two (or more) of
them compose with patchwork's `/` (stack) and `|` (side-by-side) operators just
like any other patch — useful for eyeballing whether adding a term (here,
`Species`) actually improved the residual behaviour.

### Showing the Model Call

Set `call = TRUE` to print the model's call as a title above the plots — handy
when comparing diagnostics from several candidate models side by side. Use
`call.size` to adjust its font size:

```{r call_true}
resplot(model, call = TRUE, call.size = 8)
```

### Adjusting Text Sizes

Two further size arguments control the rest of the plot's typography:

- `label.size` — the font size of the panel tags (`A`, `B`, `C`).
- `axes.size` — the font size of the axis text and titles within each panel.

```{r sizing}
resplot(model, label.size = 14, axes.size = 8)
```

### Multi-Stratum Models

Some designs are analysed with more than one error stratum — for example, a
split-plot design fitted with `aov()` and an `Error()` term, or an ASReml-R
model with multiple residual structures. For these, a single histogram/Q-Q/fitted-values
panel can't summarise every stratum, so `resplot()` returns a
**list** of plots, one per stratum, named after the strata.

```{r split_plot_setup}
set.seed(44)
n_blocks <- 8
split_plot_dat <- data.frame(
  block = factor(rep(1:n_blocks, each = 6)),
  irrigation = factor(rep(
    rep(c("Irrigated", "Rainfed"), each = 3),
    times = n_blocks
  )),
  variety = factor(rep(c("A", "B", "C"), times = 2 * n_blocks))
)
split_plot_dat$yield <- 50 +
  5 * (split_plot_dat$irrigation == "Irrigated") +
  c(A = 0, B = 2, C = 4)[split_plot_dat$variety] +
  rnorm(nrow(split_plot_dat), sd = 3)

split_plot_model <- aov(
  yield ~ irrigation * variety + Error(block / irrigation),
  data = split_plot_dat
)
```

```{r split_plot_resplot}
split_plot_resids <- resplot(split_plot_model, shapiro = FALSE)
names(split_plot_resids)
```

Each element is a complete diagnostic panel for that stratum, and can be
inspected (or saved) individually:

```{r split_plot_print}
split_plot_resids[["block:irrigation"]]
```

The same applies to multi-stratum ASReml-R models — each residual variance
structure becomes its own named element in the returned list.

### Putting Multiple Strata on One Page

When a model has several strata, printing each one separately is tedious. Set
`onepage = TRUE` to arrange them into a grid (up to 6 strata per page, paginated
automatically if there are more), and use `onepage_cols` to control the number
of columns. With three strata, `onepage_cols = 3` puts them side by side in a
single row:

```{r onepage_example, fig.width=7, fig.height=15}
resplot(
  split_plot_model,
  shapiro = FALSE,
  onepage = TRUE,
  onepage_cols = 1
)
```

`onepage_cols` matters more once there are enough strata to need more than one
row — for a model with, say, 5 strata, `onepage_cols = 2` would arrange them as
two rows of two plus a final row of one, rather than `resplot()`'s default
3-column grid. If there are more than 6 strata in total, `resplot()` returns a
*list of pages* instead of a single combined plot — index into it
(`pages[[1]]`, `pages[[2]]`, ...) to view each page.

## Supported Model Types {#supported-model-types}

`resplot()` produces diagnostics for any model class with an
`extract_model_info()` method. At the time of writing these are: `aov`, `lm`,
`aovlist` (multi-stratum `aov`), `lme` ([nlme::lme()]), `lmerMod`/
`lmerModLmerTest` ([lme4::lmer()], `lmerTest::lmer()`, and relationship-based
`lme4breeding::lmebreed()` models), `asreml`, `mmer`/`mmes` (sommer), `art`
(ARTool), `afex_aov` (afex) and `glmmTMB` (Gaussian family only — non-Gaussian
families should use `DHARMa::simulateResiduals()` instead, since a normal Q-Q
plot isn't a valid diagnostic for them).

This is a slightly broader set than the comparison functions documented in the
[*Choosing and interpreting multiple comparisons*](choosing-multiple-comparisons.html)
vignette: `resplot()` additionally supports ARTool models and sommer's legacy
`mmer` interface, neither of which has a `predict()`-based comparison
equivalent.

See `?resplot` for the full, authoritative table, and `?extract_model_info` if
you need to add support for another model engine.

## Troubleshooting

**"`model.obj` must be a linear (mixed) model object..."** — the object you
passed isn't one of the supported classes listed above. Check `class(your_model)`
against the supported list; unsupported engines need a new
`extract_model_info.<class>()` method before `resplot()` can handle them.

**The Shapiro-Wilk panel is missing** — either you set `shapiro = FALSE`, or
your model has 5,000 or more residuals, in which case the test is always
skipped (see [*D — Shapiro-Wilk Test*](#d-shapiro-wilk-test)).

**Warnings about non-finite or missing values being removed** — this usually
means some residuals or fitted values are `NA` (for example, from a
rank-deficient fit or missing data left in the model frame). Inspect
`residuals(your_model)` and `fitted(your_model)` directly to find the cause
before trusting the diagnostics.

**Plots look cramped with `onepage = TRUE`** — reduce `onepage_cols`, or omit
`onepage` and review the strata one at a time instead.


## Further Reading

- `?resplot` for the complete argument and supported-model reference.
- `?extract_model_info` for the extension mechanism used to add new model engines.
- The [*Choosing and interpreting multiple comparisons*](choosing-multiple-comparisons.html)
  vignette, for the next step after the model passes these checks.
